Incorporating periodic variability in hidden Markov models for animal movement

### 2016-07-05 paper3.Rnw
main paper pictures are done, still need to fix the appendix pictures

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
if (!require("ggalt")) {
    cat("installing ggalt from BB repo ...\n")
    library("devtools")
    install_github("bbolker/ggalt")
    library("ggalt")
}
@

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
#L <- load("cat1allmod.RData")
allmodsumdf2 <- readRDS("cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-heterogeneous HMM","homogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(grepl("HMM",model),"no","yes")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model,linetype=model))+
    facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="# of parameters") +
    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
    geom_point(aes(size=parameters))+
    scale_linetype_manual(values=linetypev)+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin)
@

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day. \\textbf{todo: leave out hmm quad for clarity? fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such. Handle obs data as in Guelph talk. Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
simdf <- simdat %>% dplyr::select(-obs) %>%
    group_by(time) %>% summarise_each(funs(mean))
obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
    filter(!is.na(obs)) %>% group_by(time) %>%
    summarise_each(funs(mean))

avgdf2 <- melt(simdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") +
    ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)+
    geom_line(data=obsdf,
              aes(y=obs),lwd=2.5,alpha=0.3,colour="black")
@

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line()
@

>>>>

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
library("RColorBrewer")
if (!require("ggalt")) {
    cat("installing ggalt from BB repo ...\n")
    library("devtools")
    install_github("bbolker/ggalt")
    library("ggalt")
}
# library("ggalt")
@

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
#L <- load("cat1allmod.RData")
allmodsumdf2 <- readRDS("summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-heterogeneous HMM","homogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(grepl("HMM",model),"no","yes")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                    , y=deltaBIC
                                    , colour=model
                                    , linetype=model))
  + facet_wrap(~ type,ncol=4)
  + scale_size_continuous( name="# of parameters")
  + scale_x_continuous(expand=c(0,0.5),breaks=2:7)
  + geom_point(aes(size=parameters))
  + scale_linetype_manual(values=linetypev)
  + scale_colour_brewer(palette="Dark2")
  + labs(y=expression(paste(Delta, "BIC"))
         , x="Number of States") 
  + geom_line() 
  + zmargin)
@

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day. \\textbf{todo: leave out hmm quad for clarity? fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such. Handle obs data as in Guelph talk. Make FMM lines dashed use manual linetype scale)})",fig.width=8>>=
# simdf <- simdat %>% dplyr::select(-obs) %>%
#     group_by(time) %>% summarise_each(funs(mean))
# obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
#     filter(!is.na(obs)) %>% group_by(time) %>%
#     summarise_each(funs(mean))
# 
# avgdf2 <- melt(simdf,'time')
avgdf <- readRDS("summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs")
mnames2 <- c("fmm-homogeneous  (n=5)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=6)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=5)"
             , "sin (n=5)"
             , "observed")

combdf <- (rbind(avgdf,obsdf)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- rep(c(2,1),c(2,6))
colvec <- c(brewer.pal(7,"Dark2"),"black")
lwdvec <- rep(c(1,4),c(7,1))
alphavec <- rep(c(1,0.2),c(7,1))
(ggplot(combdf,aes(x=time,y=steplen,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Time of Day") 
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
)
@

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
# temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)

acfdf <- (readRDS("summary_stats/cat.1.df.RDS")$acfdf 
          %>% mutate(model = .id)
)
(ggplot(acfdf,aes(x=lag,y=ACF,colour=model,lwd=model,alpha=model,linetype=model)) 
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_color_manual(values=colvec)
  + scale_alpha_manual(values=alphavec)
  + scale_size_manual(values=lwdvec)
  + scale_linetype_manual(values=linetypev)
)
@
